---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r, prepare}
  setwd("c:/coursera/reproduce/")
  unzip("repdata-data-activity.zip")
  suppressPackageStartupMessages(library(data.table))
  library(ggplot2)
  activity <- data.table(read.csv("activity.csv",colClasses=c("numeric","Date","numeric")))
```



## What is mean total number of steps taken per day?
```{r, q1, warnings = FALSE}
   
   daily <- activity[, list(sum.steps=sum(steps, na.rm=TRUE)), by=date]
   
    ggplot(daily, aes(x=date, y=sum.steps)) + 
          geom_histogram(stat="identity") +
          xlab("Steps (Total number)")+
          ggtitle("Total steps per day")
      
   mean <- as.character(daily[, round(mean(sum.steps, na.rm=T),2)])
   median <- as.character(daily[, median(sum.steps, na.rm=T)])
```

The mean total number of steps taken per day is  **`r mean`** and the median is **`r median`**.

## What is the average daily activity pattern?
```{r, q2}
inter <- activity[, list(mean.steps=mean(steps, na.rm=TRUE)), by=interval]

ggplot(inter,aes(x=interval,y=mean.steps))+geom_line()+
  ggtitle("Average steps per each 5-min interval")+
  ylab("Mean steps")

```

```{r}
a <-inter[inter[,which.max(mean.steps)]]$interval
b <-inter[inter[,which.max(mean.steps)]]$mean.steps
```
The 5-minute interval with the highest number of steps taken is `r a` with average of `r b` steps taken.

## Imputing missing values
```{r}
(miss <- sum(!complete.cases(activity)))

```

In the data `r miss` rows have missing values. 
The strategy selected to impute missing values is to use the average for this interval across all days. 

```{r}
setkey(inter, interval) ;setkey(activity, interval)
imputed <- inter[activity]
invisible(imputed[is.na(steps), steps := mean.steps])
invisible(imputed[, mean.steps :=NULL])
```




```{r}
   daily2 <- imputed[, list(sum.steps=sum(steps)), by=date]
      
   ggplot(daily2, aes(x=date, y=sum.steps, na.rm=TRUE)) + 
          geom_histogram(stat="identity") +
          xlab("Steps (Total number)")+
          ggtitle("Total steps per day after imputing")
      
   mean2 <-   as.character(daily2[, round(mean(sum.steps, na.rm=T),  2)])
   median2 <- as.character(daily2[, round(median(sum.steps, na.rm=T),2)])
```

After imputing the mean total number of steps taken per day is  **`r mean2`** (compared to **`r mean`** before imputing ) and the median is **`r median2`** (compared to **`r median`** before imputing ).


## Are there differences in activity patterns between weekdays and weekends?

```{r}
invisible(imputed[, wd :=factor(ifelse(weekdays(date) %in% c("Saturday", "Sunday"), "weekend",  "weekday"))])

inter2 <- imputed[, list(mean.steps=mean(steps, na.rm=TRUE)), by=list(wd, interval)]



ggplot(inter2,aes(x=interval,y=mean.steps))+geom_line()+
  ggtitle("Average steps per each 5-min interval after imputing")+
  ylab("Mean steps") +
  facet_grid(wd ~ .)


ggplot(inter2,aes(x=interval,y=mean.steps, group=wd, color=wd))+geom_line()+
  ggtitle("Average steps per each 5-min interval after imputing (alternative view)")  +
  ylab("Mean steps") 

```


The exporatory plots indicate as expected differing activity patterns. 
During weekdays activity is hgher in the earlier part of the days (walking to the office? or jogging before work?) and then sharply decreases (maybe an office job?). 
In weekends the activity is spead more evenly. That makes sense.

